Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2021-12-26 doc/JFLA2022/Slides/Slides.tex
small changes

\definecolor{plugincolor}{rgb}{0.6,0.0,0.6}
\definecolor{mathcolor}{rgb}{0.0,0.0,0.6}
\definecolor{coqstylecolor}{rgb}{0.0,0.0,01.0}
\definecolor{lightblue}{rgb}{0.2,0.2,1.0}
\definecolor{metavarcolor}{rgb}{0.5,0.0,1.0}
\definecolor{darkgreen}{rgb}{0.1,0.7,0.1}
\definecolor{answercolor}{rgb}{.08,.15,.8}
\definecolor{normalcolor}{rgb}{0.0,0.0,0.0}
\definecolor{exbluecolor}{rgb}{0.1,0.1,0.9}
\definecolor{dontlookcolor}{rgb}{0.5,0.5,0.5}
\definecolor{termcolor}{rgb}{0.0,0.1,0.9}
\definecolor{lookcolor}{rgb}{0.5,0.1,0.0}
\definecolor{prooftermcolor}{rgb}{0.3,0.1,1.0}
\definecolor{typecolor}{rgb}{1.0,0.6,0.0}
\definecolor{taccolor}{rgb}{0.70,0.10,0.0}
\definecolor{pink}{rgb}{0.8,0.6,0.6}
\definecolor{darkmagenta}{rgb}{0.4,0.0,0.6}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}

@@

\section{Introduction}

\begin{frame}
  \frametitle{Coq-community}
  \begin{block}{Objectives}
    \begin{itemize}
    \item \TODO{}{}
    \item\TODO{}{}
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{Yet another book on \coq?}
  \begin{block}{Kinds of books (existing)}
    \TODO{}{}
  \end{block}
\end{frame}

@@

    \begin{itemize}
    \item   We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$\omega^{\omega^2+1}+1$}.
    \item A round (at time $t\in\mathbb{N}$) of a battle transforms an hydra associated with \textcolor{mathcolor}{$\alpha$} into the hydra associated with \textcolor{mathcolor}{$\canonseq{\alpha}{t+1}$}, the $t+1$-st element of the \emph{canonical sequence} of \textcolor{mathcolor}{$\alpha$}.
    \item We want to ``compute'' the length of sequences of the form 
      \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.

    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Sequences associated with hydra battles are studied
      as \emph{$\alpha$-large sets of natural numbers}  in Ketonen \& Solovay's article
      \emph{Rapidly Growing {R}amsey Functions (1981)}.

    \item Let \textcolor{mathcolor}{$0<\alpha<\epsilon_0$}, and \textcolor{mathcolor}{$i\in\mathbb{N}$}.
      The least index \textcolor{mathcolor}{$k$} such that \textcolor{mathcolor}{$\alpha_k=0$} is  \textcolor{mathcolor}{$H'_\alpha(i)$} where \textcolor{mathcolor}{$H'_\alpha$} is defined by transfinite induction over \textcolor{mathcolor}{$\alpha$}.

@@

\begin{frame}
  \begin{block}{}
    The functions $H'_\alpha$ are ``implemented'' with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

@@

  \begin{block}{}
    Using combinatorial lemmas by  [K. \& S.], we prove (by transfinite induction on \textcolor{mathcolor}{$\alpha$}):
    \begin{quote}
      \begin{itemize}
      \item \textcolor{mathcolor}{$H'_\alpha$} is strictly monotonous
      \item \textcolor{mathcolor}{$\forall i, i \leq H'_\alpha(i)$}
      \item   If \textcolor{mathcolor}{$\beta<\alpha$}, then \textcolor{mathcolor}{$H'_\alpha(i)>H'_\beta(i)$} for all but a finite number of $i$ (we say that \textcolor{mathcolor}{$H'_\alpha$}
        \emph{dominates} \textcolor{mathcolor}{$H_\beta$}).
      \end{itemize}
    \end{quote}

@@

@@

  \begin{block}{To do ?}
    Define a language for
    \begin{itemize}
    \item Searching sentences in any \coq file,
    \item Applying filters in order to generate as many snippets as needed.
    \end{itemize}
  \end{block}
\end{frame}

>>>>

\definecolor{orange}{rgb}{1.0,0.6,0.0}
\definecolor{plugincolor}{rgb}{0.6,0.0,0.6}
\definecolor{mathcolor}{rgb}{0.0,0.0,0.6}
\definecolor{coqstylecolor}{rgb}{0.0,0.0,01.0}
\definecolor{lightblue}{rgb}{0.2,0.2,1.0}
\definecolor{metavarcolor}{rgb}{0.5,0.0,1.0}
\definecolor{darkgreen}{rgb}{0.1,0.7,0.1}
\definecolor{answercolor}{rgb}{.08,.15,.8}
\definecolor{normalcolor}{rgb}{0.0,0.0,0.0}
\definecolor{exbluecolor}{rgb}{0.1,0.1,0.9}
\definecolor{dontlookcolor}{rgb}{0.5,0.5,0.5}
\definecolor{termcolor}{rgb}{0.0,0.1,0.9}
\definecolor{lookcolor}{rgb}{0.5,0.1,0.0}
\definecolor{prooftermcolor}{rgb}{0.3,0.1,1.0}
\definecolor{typecolor}{rgb}{1.0,0.6,0.0}
\definecolor{taccolor}{rgb}{0.70,0.10,0.0}
\definecolor{pink}{rgb}{0.8,0.6,0.6}
\definecolor{darkmagenta}{rgb}{0.4,0.0,0.6}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}

@@

\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{Introduction}
  \TODO{}{Develop the abstract and background of the paper}
\end{frame}

@@

\begin{frame}
  %\frametitle{Yet another book on \coq?}
  \begin{table}
  \centering
  \footnotesize
  \begin{tabular}{lll}
  \hline
  \textbf{Title} & \textbf{Year} & \textbf{Category}\\
    \hline
   \coq reference manual &  $\geq 1989$ & \\
  Coq'Art & 2004 & Traditional\\
  Software Foundations & 2007 & Executable\\
  Certified Programming with Dependent Types & 2011 &  Executable\\
  Program Logics for Certified Compilers & 2014 & Traditional\\
  Programs and Proofs & 2014 & Executable\\
  Formal Reasoning About Programs& 2017 & Traditional\\
  Computer Arithmetic and Formal Proofs & 2017 & Traditional\\
    Mathematical Components & 2018 & Traditional\\
  \hline
  \end{tabular}
  \caption{Coq book categorization.}
  \label{tbl:books}
\end{table}

@@

    \begin{itemize}
    \item   We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$\omega^{\omega^2+1}+1$}.
    \item A round (at time $t\in\mathbb{N}$) of a battle transforms an hydra associated with \textcolor{mathcolor}{$\alpha$} into the hydra associated with \textcolor{mathcolor}{$\canonseq{\alpha}{t+1}$}, the $t+1$-st element of the \emph{canonical sequence} of \textcolor{mathcolor}{$\alpha$}.
    \item We want to ``compute'' the length of sequences of the form 
      \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.
      \item Such sequences are studied
      as \emph{$\alpha$-large sets of natural numbers}  in Ketonen \& Solovay's article
      \emph{Rapidly Growing {R}amsey Functions (1981)}.
    \end{itemize}
  \end{block}
\end{frame}

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Let \textcolor{mathcolor}{$0<\alpha<\epsilon_0$}, and \textcolor{mathcolor}{$i\in\mathbb{N}$}.
      The least index \textcolor{mathcolor}{$k$} such that \textcolor{mathcolor}{$\alpha_k=0$} is  \textcolor{mathcolor}{$H'_\alpha(i)$} where \textcolor{mathcolor}{$H'_\alpha$} is defined by transfinite induction over \textcolor{mathcolor}{$\alpha$}.

@@

\begin{frame}
  \begin{block}{}
    The functions \textcolor{mathcolor}{$H'_\alpha$} are implemented with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

@@

  \begin{block}{}
    Using combinatorial lemmas by  K. \& S., we prove (by mutual  transfinite induction on \textcolor{mathcolor}{$\alpha$}):
    \begin{quote}
      \begin{itemize}
      \item \textcolor{mathcolor}{$H'_\alpha$} is strictly monotonous
      \item \textcolor{mathcolor}{$\forall i, i \leq H'_\alpha(i)$}
      \item   If \textcolor{mathcolor}{$\beta<\alpha$}, then \textcolor{mathcolor}{$H'_\alpha(i)>H'_\beta(i)$} for all but a finite number of $i$ (we say that \textcolor{mathcolor}{$H'_\alpha$}
        \emph{dominates} \textcolor{mathcolor}{$H_\beta$}).
      \end{itemize}
    \end{quote}

@@

  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Project Organization}
  \begin{frame}
    \frametitle{Project  Organization}
    \begin{itemize}
    \item Documentation with \alectr
     \item Integration with \community
     \item Package dependencies
       \item Continuous integration
    \end{itemize}
  \end{frame}

  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
  \begin{frame}
    \frametitle{\community}
    \begin{block}{}
      \community is an informal organization that aims to maintain interesting open source \coq projects  andfacilitate collaboration among Coq users on documentation, tooling, etc.
      \begin{itemize}
      \item Created in 2018,  run by volunteer Coq users on GitHub
      \item At present, over than 50 projects, maintained by over than 30 volunteers
      \item Some packages were unmaintained before their integration to \community. According to their scientific or historical interest, we decided to revive them.
        \item After their transfer, packages may be the object of large changes  and refactorings.
      \end{itemize}
    \end{block}
  \end{frame}

@@

  \begin{block}{To do ?}
    Define a language for
    \begin{itemize}
    \item Searching sentences in any \coq file,
    \item Applying filters in order to generate as many snippets as needed.
    \end{itemize}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Package organization and dependencies}

\begin{figure}[ht]
\centering
\scalebox{0.7}{
\begin{tikzpicture}[thick]

\begin{scope}[xshift=-6cm]
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm] (goedel) { Goedel };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,left of=goedel, node distance=2.2cm] (pock) { Pocklington };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=goedel, node distance=2.5cm] (additions) { Additions };
\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=additions, node distance=2cm] (cantor) { Cantor };

\node[rectangle, dotted,draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=cantor, node distance=3.2cm] (cats) { Categories in ZFC };
\end{scope}

\begin{scope}[yshift=-1cm,xshift=0.45cm]
  \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm] (paramcoq) { Paramcoq };
\node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm,right of=paramcoq, node distance=2.5cm] (mathcomp) { MathComp };
\node[rectangle, draw=black,minimum height=0.5cm,minimum width=1.5cm,node distance=2cm,left of=paramcoq, node distance=4.5cm] (equations) { Equations };
\end{scope}

\begin{scope}[yshift=-2cm,xshift=-4cm]
\node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm] (hydras) { Hydra-battles };
\node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm,left of=hydras, node distance=3.5cm] (pockcc) { Pocklington };
\node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm,right of=hydras, node distance=2.5cm] (chains) { Addition-chains };
\node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm,right of=chains, node distance=2.2cm] (gaia) { Gaia };
\end{scope}

\begin{scope}[yshift=-3cm,xshift=-6cm]
 \node[rectangle, draw=black,fill=blue!30,minimum height=0.5cm,minimum width=1.5cm] (goedelcc) { Goedel };
 \node[rectangle, draw=black,fill=orange!80,minimum height=0.5cm,minimum width=1.5cm,right of=goedelcc, node distance=3cm] (gaiahydras) { Gaia-hydras };
\end{scope}

\draw[lookcolor,->,dashed] (additions.south) -- (chains.north) ;
\draw[lookcolor,->,dashed] (goedel) -- (goedelcc) ;
\draw[lookcolor,->,dashed] (goedel) -- (hydras.north west) ;
\draw[lookcolor,->,dashed] (cantor) -- (hydras.north east) ;
\draw[lookcolor,->,dashed] (cantor.south) -- (gaia.north west) ;
\draw[lookcolor,->,dashed] (pock) -- (pockcc) ;
\draw[lookcolor,->,dashed] ([xshift=2.1cm]cats) -- (gaia) ;

\draw[->] (hydras) -- (gaiahydras) ;
\draw[->] (hydras.south west) -- (goedelcc) ;
\draw[->] (gaia.south west) -- (gaiahydras.east) ;

\draw[->] (equations) -- (hydras) ;
\draw[->] (mathcomp) -- (gaia) ;
\draw[->] (mathcomp) -- (chains) ;
\draw[->] (paramcoq.south) -- (chains) ;
\draw[->] (pockcc) -- (goedelcc) ;
\end{tikzpicture}
}
\caption{\textcolor{lookcolor}{Genealogy} and dependencies for \Hydras packages.}
 %Dotted boxes represent historical Coq contributions, while regular boxes represent maintained \coq packages. Orange packages are maintained in the \Hydras GitHub repository, while blue packages are maintained in other \community repositories. Dotted lines represent \coq code ancestry, while regular lines represent direct code dependencies.}
  \label{fig:genealogy}
\end{figure}

%%%% TODO : Resize  !
{\scriptsize
\begin{table}[ht]
\centering
\scriptsize
\begin{tabular}{llllrr}
\hline
\textbf{Package name} &  \textbf{Version} & \textbf{Spec LOC} & \textbf{Proof LOC}\\
\hline
Pocklington & \texttt{8.12.0} & 825 & 3,798 \\
Goedel  & \texttt{8.13.0} & 2,799 & 10,762 \\
Gaia &  \texttt{1.12} & 28,850 & 124,839 \\
Hydra-battles & \texttt{0.5} & 14,368 & 48,385 \\
Addition-chains & \texttt{0.5} & 1,961 & 2,210 \\
Gaia-hydras &  \texttt{0.5} & 81 & 177 \\
\hline
\end{tabular}
\caption{\scriptsize Current numbers of lines of code for packages in the \Hydras galaxy.}
\label{tbl:loc}
\end{table}}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
